\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-20 doc/incertae_sedis.tex
Finish revising section 2.7

Note that this example also exposes a lack of consistency in
Equation (\ref{eq:exsetformonea}) that rears its head when attempting to
iteratively revise placement of \emph{incertae sedis} taxa in the taxonomy
on successive supertree builds.
% without retention of additional information.

Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
    rooted split: $W\, V1 \mid \bullet\,V2$.
If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
While the topology for the final tree in either order of analyses would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}b, the decision about whether or not to name $V$ would differ.
Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
As expected,  the definition of $WX$ has been altered to reflect the
    fact $T_1$ placed $V$ inside $WX$
    when we encode the revised taxonomy.
Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.

@@

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[maximally scored]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill{}
\subfloat[less informative]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill
\subfloat[from revised taxonomy]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & 1 \\
$V2$ & 1 & 1 \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Panels (a) and (b) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}c. }
\end{figure}

>>>>

Note that equation (\ref{eq:exsetformonea}) allows \emph{incertae sedis}
taxa to intrude only into siblings in the original taxonomy, and not
into taxa that become siblings through solution to the supertree
problem. This can lead to cases where computing a supertree from two
input trees $T_1$ and $T_2$ and an original taxonomy $\taxonomy$ can
lead to a different outcome than if we compute a supertree from $T_1$
and $\taxonomy$, and then compute a supertree from $T_2$ and a revised
taxonomy $\taxonomy^\prime$. Consider a case in which the taxonomy
$\taxonomy$ is given in Figure \ref{fig:Placing-one-incertae}a (with
matrix representation given by \ref{fig:Placing-one-incertae-MRP}a)
and $T_1$ and $T_2$ each provide the single rooted splits $V1\,X \mid \bullet\,U$ and $W\, V1 \mid
\bullet\,V2$ respectively. 
Running the supertree pipeline on $T_1$  and $\taxonomy$ yields the tree in
Figure \ref{fig:Placing-one-incertae}b.  We use this tree as the revised
taxonomy $\taxonomy^\prime$, and its matrix representation appears in
Figure \ref{fig:Placing-one-incertae-MRP}c.  Running the supertree
pipeline on $T_2$  and $\taxonomy^\prime$ then
results in the tree in Figure \ref{fig:Placing-one-incertae}c.

One might therefore hope that running the supertree pipeline on $T_1$,
$T_2$, and $\taxonomy$ simultaneously would yield the same result.
While the topology for the final tree would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}c, the decision about whether
    or not to name $V$ differs.  As expected,  the definition of $WX$
    in $\taxonomy^\prime$ has been altered to reflect the fact $T_1$
    placed $V$ inside $WX$.  Perhaps unexpected is the fact that
    storing the revised taxonomy with an \emph{incertae sedis}
    annotation and then reprocessing the taxonomy using our rooted
    partial splits approach would also shift the definition of $V$
    from a taxon that must exclude $W$ to one that may include $W$. 
    This results from the fact that equation (\ref{eq:exsetformonea})
    only allows \emph{incertae sedis} taxa to intrude into taxa that
    are their
    sisters in the taxonomy.  Outcomes differ because $W$ is a sister of $V$ in
    $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}b
    allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
    which solves the seeming inconsistency.  However, this has some
    surprising side-effects as mentioned above, and is therefore not
    the approach that we present in this paper.

@@

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[maximally scored]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill{}
\subfloat[less informative]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}\hfill
\subfloat[from revised taxonomy]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & 1 \\
$V2$ & 1 & 1 \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
(root) & 0 & 0 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Panels (a) and (b) show two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$. Panel (c) shows the information encoded by the tree in \ref{fig:Placing-one-incertae}b. }
\end{figure}
